---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = T, warning=F, message=F)
```

# Bayesian Analysis of Macroevolutionary Mixtures
**Author**: Léo-Paul Dagallier    
**Last update**: `r format(Sys.Date())`

***

## Linux installation

```{bash, eval = F, echo = T}
cd <path to program directory>
git clone https://github.com/macroevolution/bamm.git
cd bamm
mkdir build
cd build
cmake ..
make -j
sudo make install
```

## BAMM analysis
### Input data
Prepare the path variables:

- in bash:
```{bash, eval = F}
path_to_data="data";
path_to_output="outputs";
cd $path_to_output
mkdir BAMM
```

- in R:
```{r}
path_to_tree = c("data/name_MCC_monodoreae3_monod_pruned.newick")
path_to_output = c("outputs/BAMM/")
path_to_figures = c("figures/")
data_suffix <- "monodoreae3"
```

### Check the prerequesities of the tree

The tree needs to be ultrametric, fully bifurcating (no polytomy) and minimum branch length should be \> 0.

```{r, warning=FALSE, eval = T}
library(ape)
v <- read.tree(path_to_tree)
is.ultrametric(v)
is.binary(v)
# Now to check min branch length:
min(v$edge.length)
```

### Prepare the configuration file

Prepare manually the configuration file in the `data/BAMM` directory.

#### Prepare using BAMMtools
```{r}
library(BAMMtools)
setBAMMpriors(v, outfile = paste0(path_to_output, "myPriors.txt"))
```
Copy the values from myPriors.txt into the bamm_config_FINAL file.

### Run the analysis
In bash (not R).
```{bash, eval = F}
cd $path_to_output/BAMM
bamm -c $path_to_data/bamm_config_FINAL
```

### Assess MCMC convergence

Compute effective sample size after discarding the burnin:

```{r}
mcmcout <- read.csv(paste0(path_to_output, "mcmc_out.txt"), header=T)
burnstart <- floor(0.1 * nrow(mcmcout))
postburn <- mcmcout[burnstart:nrow(mcmcout), ]
```

```{r MCMC-convergence-ESS}
library(coda)
paste0("ESS for the log Likelihood: ", effectiveSize(postburn$logLik))
paste0("ESS for the log Prior: ",effectiveSize(postburn$logPrior))
paste0("ESS for the number of shifts: ",effectiveSize(postburn$N_shifts))
```

Visual assessment of MCMC convergence of the BAMM run:

```{r MCMC-convergence-traces-plots}
par(mfrow=c(2,2))
plot(mcmcout$logLik, type="l", main="a)", xlab="Generations",ylab="log Likelihood")
plot(mcmcout$N_shifts ~ mcmcout$generation, main="b)", xlab="Generations",ylab="Number of shifts")
plot(postburn$logLik, type="l", main="c)", xlab="Generations",ylab="log Likelihood")
plot(postburn$N_shifts ~ postburn$generation, main="d)", xlab="Generations",ylab="Number of shifts")
```

Save the plot in PDF:

```{r MCMC-convergence-plot-export, echo=T, include=T}
pdf(paste0(path_to_figures, "BAMM - 1. Assessing MCMC convergence of the BAMM run - ", data_suffix, ".pdf"))
par(mfrow=c(2,2))
plot(mcmcout$logLik, type="l", main="a)", xlab="Generations",ylab="log Likelihood")
plot(mcmcout$N_shifts ~ mcmcout$generation, main="b)", xlab="Generations",ylab="Number of shifts")
plot(postburn$logLik, type="l", main="c)", xlab="Generations",ylab="log Likelihood")
plot(postburn$N_shifts ~ postburn$generation, main="d)", xlab="Generations",ylab="Number of shifts")
dev.off()
```

### Getting BAMM results

```{r, echo=T, include=T}
library(BAMMtools)
tree <- read.tree(path_to_tree)
edata <- getEventData(tree, eventdata = paste0(path_to_output, "event_data.txt"), burnin=0.1)
```

## Analysis of rate shifts
### How many rate shifts
Compute the posterior probabilities of the models sampled using BAMM:

```{r}
post_probs <- table(postburn$N_shifts) / nrow(postburn)
post_probs
shift_probs <- summary(edata)
```

Posterior probability of the expected number of rate shifts.
```{r expected-n-shifts}
par(mfrow=c(1,1))
plot(post_probs, xlim=c(0,8), main="", ylab="Posterior", xlab="Number of shifts")
```

Save the plot in PDF:
```{r, echo=T, include=T}
pdf(paste0(path_to_figures, "BAMM - 2. Posterior for number of shifts - ", data_suffix, ".pdf"))
plot(post_probs, xlim=c(0,8), main="", ylab="Posterior", xlab="Number of shifts")
dev.off()
```

### Compute Bayes Factor
```{r}
bfmat <- computeBayesFactors(mcmcout, expectedNumberOfShifts=1, burnin=0.1)
bfmat[,1]
```

Here it means that the models with 1, 2 or 3 shifts are not better than the null model (the model with 0 shift).

### Mean Phylo rate plot
Speciation rate through time:
```{r speciation-rate}
# plot.bammdata(edata, tau=0.001, spex="s", legend=T, lwd=1.5, logcolor = F,  breaksmethod = "linear")
ratesHistogram(plot.bammdata(edata, tau=0.001, spex="s", legend=T, lwd=1.5), plotBrks = TRUE, xlab = "Speciation rate")
axis(side = 1, at = c(0, 0.17, 0.175, 0.180, 0.185, 0.190, 0.195, 0.20, 0.205, 0.21, 0.215, 0.3),  cex.axis = 0.75, tcl = NA, mgp = c(0,  0.25, 0))
```

Extinction rate through time:
```{r extinction-rate}
ratesHistogram(plot.bammdata(edata, tau=0.001, spex="e", legend=T, lwd=1.5), plotBrks = TRUE, xlab = "Extinction rate")
```

Net diversification rate through time:
```{r net-diversification-rate}
ratesHistogram(plot.bammdata(edata, tau=0.001, spex="netdiv", legend=T, lwd=1.5), plotBrks = TRUE, xlab = "Net diversification rate")
axis(side = 1, at = c(0, 0.15, 0.155, 0.16, 0.165, 0.17, 0.175, 0.180),  cex.axis = 0.75, tcl = NA, mgp = c(0,  0.25, 0))
```

Save the plots in PDF:
```{r, echo=T, include=T}
pdf(paste0(path_to_figures, "BAMM - 3. Speciation rate through time of Monodoreae - ", data_suffix, ".pdf"))
plot.bammdata(edata, tau=0.001, spex="s", legend=T, lwd=2, labels = T, cex = 0.45)
dev.off()
pdf(paste0(path_to_figures, "BAMM - 4. Extinction rate through time of Monodoreae - ", data_suffix, ".pdf"))
plot.bammdata(edata, tau=0.001, spex="e", legend=T, lwd=2, labels = T, cex = 0.45)
dev.off()
pdf(paste0(path_to_figures, "BAMM - 5. Net diversification rate through time of Monodoreae - ", data_suffix, ".pdf"))
plot.bammdata(edata, tau=0.001, spex="netdiv", legend=T, lwd=2, labels = T, cex = 0.45)
dev.off()
```

### Bayesian credible sets of shift configurations
```{r}
css <- credibleShiftSet(edata, expectedNumberOfShifts=1, threshold=5, set.limit = 0.95)
css$number.distinct
summary(css)
plot.credibleshiftset(css, spex="netdiv", BFcriterion=5, border=F, pal="temperature")
plot.credibleshiftset(css, spex="s", BFcriterion=5, border=F, pal="temperature")
plot.credibleshiftset(css, spex="e", BFcriterion=5, border=F, pal="temperature")
```

### Finding the single best shift configuration

```{r}
best <- getBestShiftConfiguration(edata, expectedNumberOfShifts=1)
{plot.bammdata(best, lwd = 2)
addBAMMshifts(best, cex=2.5)}
```

